---
title: "Removing batch effect filtering by variance and running ComBat"
author: "Felipe Gajardo"
date: "5/12/17 / Edited: 25/3/20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Abundance filtering

```{r}

library(limma)
library(sva)

# expresion data
d=read.table("004_original_marray_annotated_sample_log2.csv",sep=',',h=T)

# initial number of probes
dim(d)

# Normalization. (skipped since data is already normalised)
#d=normalizeBetweenArrays(d,method="scale")

#Filtering by total abundance
d.sum=apply(d,1,sum)
#hist(d.sum,breaks=100, col="#2b83ba")
#summary(d.sum)
d=d[names(which(d.sum>16.26)),] #abundance above lower quartile
#d.sum.after=apply(d,1,sum)
#hist(d.sum.after,breaks=100,col="#2b83ba")

# How many probes are kept?
dim(d)
```

# Removing batch with ComBat

First we set models for batch correction.

```{r}
#setting models. We called conditions as c("A_wt","B_mut") for obtaining the inverse sign in fold change values.
phenotypes <- data.frame(sample_name = colnames(d), treatment = rep(c("A_wt","B_mut"), 3), extraction_day = rep(c(1:3), each = 2))
phenotypes

mod=model.matrix(~as.factor(treatment),data=phenotypes)
mod0=model.matrix(~1,data=phenotypes)

#run combat
d.combat=ComBat(dat=as.matrix(d),batch=phenotypes$extraction_day,mod=mod0)

#these are pval and qval returned by of combat (not DE)
combat.pval= f.pvalue(as.matrix(d.combat),mod,mod0)
combat.qval=p.adjust(combat.pval,method="BH")
```

**This is how filtered original data looks like before batch correction**

```{r}
#before combat
plotMDS(d)

plotDensities(d)

```


**This is how it looks like after batch correction**


```{r}
#after combat
plotMDS(d.combat)

plotDensities(d.combat)
```



# Differential expression analysis

We used `limma` package for microarray differential expression analysis. `limma` was provided with the model obtained for batch corrected data with `combat`

```{r}
# load batch corrected data in limma 
fit=lmFit(d.combat, mod)
fit=eBayes(fit)
de=topTable(fit,coef="as.factor(treatment)B_mut",number=length(rownames(d.combat)))

head(de)

```

And DE analysis is done. Now we get some stats related to number of DE genes, Adj. p-value, etc.


# Getting stats

Adj. P-value threashold considered for counting DE probes was `0.05`.

```{r}
#threasold adj. pval
threshold=0.05

#Plot Adjusted P-values
hist(de$adj.P.Val,breaks=100,col="#fdae61")
summary(de$adj.P.Val)

#N Adj. P-value below threshold
n.sig=length(which(de$adj.P.Val < threshold))
length(which(de$adj.P.Val < threshold))

#write all(after filtering) probes to a file
write.table(de,file="all.tsv",sep='\t',quote=F)

#write significant probes to a file
out=de[de$adj.P.Val < threshold,]
write.table(out,file="significant.tsv",sep='\t',quote=F)
```

Finally save session

```{r}
#just for saving the workspace
save.image(".RData")
```
